Master Thesis

### 2013-06-26 implementation.tex
Proofreading for publication

We consider the above capabilities vital for doing any involved grammar
engineering in the framework of ACGs. However, there are some
limitations which make it not sufficient for our needs. The foremost of
those is that the signatures and lexicons are all explicitly realized in
(primary) memory and the toolkit expects them to be given directly in a
file. The grammars that we would like to develop will cover an
exhaustive list of wordforms and the size of our grammar definitions
would therefore grow proportionately with the size of our dictionary.

@@

Furthermore, we could pick a small representative fragment of the
dictionary and test only on that fragment during the development of our
grammar, to reduce the time it takes to both generate the direct
representation of the grammar and the time it takes to load it in the
toolkit. This would have made developing the grammar already
tenable.

However, in our approach, we have opted to spend more time on tool
development to ease subsequent work and we have developed a system which
makes defining and experimenting with lexicalized ACGs easier. The
metagrammars are defined as relational procedures which link the items
of the lexical database to elements of signatures and lexicons. The
grammars can be redefined without having to reload the lexical database
or any file of similar scale. Type inference and lexicon application
with $\beta$-normalization are provided as relations which can be used
interactively or as part of a larger program. En example of such a
larger program might be a routine to check the validity of an ACG or to
test some of its properties.

@@

There are several reasons why we might want to choose a relational
programming system for our implementation. First off, the problems which
constitute the algorithmic core of our program are often mere textbook
examples of how to use a relational programming system. This goes for
type inference/checking/inhabitation, which are all implemented by a
single relation which is often used as a kind of ``Hello World'' program
in the world of relational programming. Similarly, implementing
$\beta$-reduction cleanly (without having to handle variable clashes by
generating new variables and $\alpha$-converting) is a motivational
example for introducing \emph{nominal logic programming}. To reach
feature parity with the ACG toolkit, one would only need to implement
the mapping of a lexicon over a term, which is a simple functor.

@@

The third system was XMG, a language for writing metagrammars, which is
another goal of our system, the one which distinguishes it from the
original ACG toolkit. We believe that using a declarative programming
system such as miniKanren with the ability to extend it using a
practical functional programming language presents an empowering way of
defining metagrammars. This allows the grammar designer to compose her
grammar in any way she sees fit and to use the metaprogramming
facilities of the underlying programming language to transcribe the
grammar in a way that makes the most sense in her case.

@@

There are two programming languages which play host to a substantial
implementation of miniKanren (meaning one that has extensible
unification and/or constraints and that includes support for nominal
logic), Racket\footnote{\url{https://github.com/calvis/cKanren}} and
Clojure\footnote{\url{https://github.com/clojure/core.logic}}. The two
languages are very similar to each other and therefore the choice is of
no great significance. An appealing feature of Clojure is that it uses
abstract collection types pervasively instead of coupling the standard
operations to specific data structures such as lists and cons cells and
that it follows other similarly enlightened design decisions. Racket on
the other hand has had more time to mature, has better metaprogramming
facilities and is more recognized in academia. In developing our system,
we have settled for Clojure. One of the advantages of doing so is that
the Clojure implementation of miniKanren, dubbed core.logic, has
inspired the software community and the project has attracted
contributors that submit numerous issue reports and
patches.\footnote{\url{http://dev.clojure.org/jira/browse/LOGIC}}

@@

In~\ref{ssec:sig}, we formally presented signatures as triples
$\mathopen{<}A, C, \tau\mathclose{>}$. The set $C$ can be easily
reconstructed as the domain of $\tau$ and the set $A$ is just the set of
all atomic types occurring in the range of $\tau$. Therefore, to define
the signature $\mathopen{<}A, C, \tau\mathclose{>}$, it is enough to
provide the function $\tau$, which in its set-theoretical realization is
just a set of pairs. In relational programming, a specific set can be
easily encoded as a unary relation so a unary will serve as our
representation of a signature and defining a signature will be a matter
of constructing this relation.

@@

The system takes care of wiring up the relations and assembling the
signature. The grammar designer's duty is to provide a 4-ary relation
that links the four parts of a lexical constant: the wordform, the
hypertag, the specifier and the type. This relation can be thought of as
a multi-valued ``function'' that receives as input a wordform and a
hypertag (that is, a single lexical entry) and outputs (possibly
multiple) pairs of specifiers and types.

Since we are working in a functional programming language that supports
manipulation of higher-order functions, we can provide helper functions
which construct these 4-ary relations for the grammar designer in some
simple cases. Our library thus provides functions like
\texttt{unitypedr}, which expects a type and returns the 4-ary relation
that assigns to every lexical item a single constant whose specifier is
$nil$ and whose type is the supplied type, or \texttt{ht->typer}, which
expects a mapping from hypertag patterns to types and returns the 4-ary
relation which tries to map the hypertag against the patterns and yields
constants with the corresponding types and $nil$ specifiers. When other
patterns emerge, the grammar designer is completely free to implement
her new abstractions using the full power of the functional programming
language.

@@

In the above paragraphs, we have been referring to the constants that we
were defining as \emph{lexical constants}. That was to distinguish them
from the \emph{non-lexical constants} that we will introduce now. The
reason for having non-lexical constants is that not every constant of
every signature is linked to (generated by) some item of our lexical
database. Consider, for example, the empty string $\epsilon$ and the
string concatenation operator $+$ of the $\Sigma_{String}$ signature we
gave in~\ref{ssec:example-sig} or the quantifiers and logical
connectives of the $\Sigma_{Sem}$ signature in~\ref{ssec:acg-sem}.
These constants are not associated to any lexical entry or specifier and
therefore their representations must be different. See
Figure~\ref{fig:nonlex-const} for an example of the representation we
use for non-lexicalized constants.

@@

Signatures of non-lexical constants can be simply defined by a mapping
from the constant names to their types as
in~\ref{fig:nonlex-sig-impl}. Such signatures will generally not be
sufficient by themselves and will need to be complemented by other
signatures which contain lexical constants. Since we represent
signatures as unary relations, combining them to produce their unions or
intersections is just a matter of taking their disjunctions
(\texttt{ors}) or conjunctions (\texttt{ands}), respectively.

@@

With this, we have enough to fully define all the signatures of
section~\ref{sec:acg} in a lexicalized manner, see
Figure~\ref{fig:example-sig-impl}.

@@

In our relational framework, a mapping can be represented most directly
as a binary relation and that is the representation that we use in our
system. In the previous subsection, we talked a lot about how individual
entries in the lexical database generate the items of a
signature. Solving the problem for signatures also solved it for
lexicons, since a lexicon contains exactly one object term for every
constant in its abstract signature. Therefore, a pair of an abstract
constant and an object term belonging to some lexicon is generated by
the same lexical entry that generated the abstract constant.

Our representation of lexical constants was purposefully engineered so
as to make the lexical entry directly accessible to the lexicon
implementation. Since the lexicon is directly implemented by the grammar
designer as a binary relation between constants and terms, the grammar
designer can inspect the abstract constant and directly access the
hypertag and wordform of the lexical entry that generated it, the
specifier that was assigned to it and the type of the constant and then
express the object term assigned to the constant using any combination
of these.

@@

The above is already enough to define any lexicon. However, our toolkit
presents convenience facilities that capture common forms of
lexicons. With these, we will be able to define lexicalized versions of
the lexicons of section~\ref{sec:acg} in a concise point-free
style. Furthermore, the grammar designer is not limited to the convenience
facilities we have provided and is completely free to use the same
programming language we have used to build on our abstractions and build
her own right within her grammar definition.

@@

When mapping lexical constants by a lexicon, we usually produce object
terms which contain constants of the object signature that are generated
by the same lexical entry as the abstract constant
(e.g. $\mathcal{L}_{syntax}(\synt{love}) = \lambda^{\circ} x y.\ x +
loves + y$). Managing this involves some tedious boilerplate that can be
avoided by another convenience facility. First, we define a lexicon that
produces the desired object term but with the object constant abstracted
out (e.g. $\lambda^{\circ} c x y.\ x + c + y$). Then, we apply the
\texttt{lexicalizer} function to the object signature and to this
lexicon to get the desired lexicon which fills in the correct object
constants.

It is also useful to be able to easily insert the non-lexical constants
of the object signature in the object terms we are giving when defining
a lexicon. This service is provided by the anaphoric macro
\texttt{with-sig-consts} which introduces all of the non-lexical
constants of a signature into scope for easy use inside of lexicon
definitions.

Finally, since lexicons are again just sets of pairs, we can take their
unions and intersections using disjunction (\texttt{orr}) and
conjunction (\texttt{andr}).

@@

The crucial property that we need to check for in signatures is that the
relation between constant identities and their types is a function. This
simply reduces to querying the system for the number of typed constants
belonging to the signature and having some specific identity. If this
number is one for all of our test cases, we have verified that (the
tested subset of) the signature assigns exactly one type to every one of
its constants.

>>>>

We consider the above capabilities vital for doing any involved grammar
engineering in the framework of ACGs. However, there are some
limitations which make it not sufficient for our needs. The foremost of
these is that the signatures and lexicons are all explicitly realized in
(primary) memory and the toolkit expects them to be given directly in a
file. The grammars that we would like to develop will cover an
exhaustive list of wordforms and the size of our grammar definitions
would therefore grow proportionately with the size of our dictionary.

@@

Furthermore, we could pick a small representative fragment of the
dictionary and test only on that fragment during the development of our
grammar, to reduce the time it takes to both generate the direct
representation of the grammar and the time it takes to load it into the
toolkit. This would have made developing the grammar already tenable.

However, in our approach, we have opted to spend more time on tool
development to ease subsequent work and we have developed a system which
makes defining and experimenting with lexicalized ACGs easier. The
metagrammars are defined as relational procedures which link the items
of the lexical database to elements of signatures and lexicons. The
grammars can be redefined without having to reload the lexical database
or any file of similar scale. Type inference and lexicon application
with $\beta$-normalization are provided as relations which can be used
interactively or as part of larger programs. En example of such a larger
program might be a routine to check the validity of an ACG or to test
some of its properties.

@@

There are several reasons why we might want to choose a relational
programming system for our implementation. First off, the problems which
constitute the algorithmic core of our program are often mere textbook
examples of how to use a relational programming system. This goes for
type inference/checking/inhabitation, which are all implemented by a
single relation which is often used as a kind of ``Hello World'' program
in the world of relational programming. Similarly, implementing
$\beta$-reduction cleanly (without having to handle variable clashes by
generating new variables and $\alpha$-converting) is a motivational
example for introducing \emph{nominal logic programming}. Finally, to
reach feature parity with the ACG toolkit, one would only need to
implement the mapping of a lexicon over a term, which is a simple
functor.

@@

The third system at the conference was XMG, a language for writing
metagrammars, which is another goal of our system, one which
distinguishes it from the original ACG toolkit. We believe that using a
declarative programming system such as miniKanren with the ability to
extend it using a practical functional programming language presents an
empowering way of defining metagrammars. This allows the grammar
designer to compose her grammar in any way she sees fit and to use the
metaprogramming facilities of the underlying programming language to
transcribe the grammar in a way that makes the most sense in her case.

@@

There are two programming languages which play host to a substantial
implementation of miniKanren (meaning one that has extensible
unification and/or constraints and that includes support for nominal
logic), Racket\footnote{\url{https://github.com/calvis/cKanren}} and
Clojure\footnote{\url{https://github.com/clojure/core.logic}}. The two
languages are very similar to each other and therefore the choice is of
no great significance. An appealing feature of Clojure is that it uses
abstract collection types pervasively instead of coupling the standard
operations to specific data structures such as lists and cons cells and
that it follows other similarly enlightened design decisions. Racket, on
the other hand, has had more time to mature, has better metaprogramming
facilities and is more recognized in academia. In developing our system,
we have settled for Clojure. One of the advantages of doing so is that
the Clojure implementation of miniKanren, dubbed core.logic, has
inspired the software community and the project has attracted
contributors that submit numerous issue reports and
patches.\footnote{\url{http://dev.clojure.org/jira/browse/LOGIC}}

@@

In~\ref{ssec:sig}, we formally presented signatures as triples
$\mathopen{<}A, C, \tau\mathclose{>}$. However, the set $C$ can be
always reconstructed as the domain of $\tau$ and the set $A$ is just the
set of all the atomic types occurring in the range of $\tau$. Therefore,
to define the signature $\mathopen{<}A, C, \tau\mathclose{>}$, it is
enough to provide the function $\tau$, which in its set-theoretical
realization is just a set of pairs. In relational programming, a
specific set can be easily encoded as a unary relation so a unary
relation will serve as our representation of a signature and defining a
signature will be a matter of constructing this relation.

@@

The system takes care of wiring up the relations and assembling the
signature that yields these structures. The grammar designer's duty is
to provide a 4-ary relation that links the four parts of a lexical
constant: the wordform, the hypertag, the specifier and the type. This
relation can be thought of as a multi-valued ``function'' that receives
as input a wordform and a hypertag (that is, a single lexical entry) and
outputs (possibly multiple) pairs of specifiers and types.

Since we are working in a functional programming language that supports
manipulation of higher-order functions, we can provide helper functions
which construct the signature relations for the grammar designer in some
simple cases. Our library thus provides functions like
\texttt{unitypedr}, which expects a type and returns a signature that
assigns to every lexical item a single constant whose specifier is $nil$
and whose type is the supplied type, or \texttt{ht->typer}, which
expects a mapping from hypertag patterns to types and returns a
signature which tries to map the hypertags of lexical items against the
patterns and yields constants with the corresponding types and $nil$
specifiers. When other patterns emerge, the grammar designer is
completely free to implement her new abstractions using the full power
of the functional programming language.

@@

In the above paragraphs, we have been referring to the constants that we
were defining as \emph{lexical constants}. That was to distinguish them
from the \emph{non-lexical constants} that we will introduce now. The
reason for having non-lexical constants is that not every constant of
every signature is linked to (generated by) some item of our lexical
database. Consider, for example, the empty string $\epsilon$ and the
string concatenation operator $+$ of the $\Sigma_{String}$ signature we
gave in~\ref{ssec:example-sig} or the quantifiers and logical
connectives of the $\Sigma_{Sem}$ signature in~\ref{ssec:acg-sem}.
These constants are not associated to any lexical entry and therefore
their representations must be different. See
Figure~\ref{fig:nonlex-const} for an example of the representation we
use for non-lexicalized constants.

@@

Signatures of non-lexical constants can be simply defined by a mapping
from the constant names to their types as in
Figure~\ref{fig:nonlex-sig-impl}. Such signatures will generally not be
sufficient by themselves and will need to be complemented by other
signatures which contain lexical constants. Since we represent
signatures as unary relations, combining them to produce their unions or
intersections is just a matter of taking their disjunctions
(\texttt{ors}) or conjunctions (\texttt{ands}), respectively.

@@

With this, we have enough tools to fully define all the signatures of
section~\ref{sec:acg} in a lexicalized manner, see
Figure~\ref{fig:example-sig-impl}.

@@

In the relational framework, a mapping can be represented most directly
as a binary relation and that is the representation of lexicons that we
use in our system. In the previous subsection, we talked a lot about how
individual entries in the lexical database generate the items of a
signature. Solving the problem for signatures also solved it for
lexicons, since a lexicon contains exactly one object term for every
constant in its abstract signature. Therefore, a pair of an abstract
constant and an object term belonging to some lexicon is generated by
the same lexical entry that generated the abstract constant and its
type.

Our representation of lexical constants was purposefully engineered so
as to make the lexical entry directly accessible to the lexicon
implementation. Since the lexicon is implemented by the grammar designer
as a binary relation between constants and terms, the grammar designer
can inspect the abstract constant and directly access the hypertag and
wordform of the lexical entry that generated it, the specifier that was
assigned to it and the type of the constant, and then express the object
term assigned to the constant using any combination of these.

@@

The above is already enough to define any lexicon. However, our toolkit
presents convenience facilities that capture common forms of
lexicons. With these, we will be able to define lexicalized versions of
the lexicons of section~\ref{sec:acg} in a concise point-free
style. Furthermore, the grammar designer is not limited to the
convenience facilities we have provided and is completely free to use
the same programming language we have used to build her own abstractions
on top of ours right within her grammar definition.

@@

When mapping lexical constants by a lexicon, we usually produce object
terms which contain constants of the object signature that are generated
by the same lexical entry as the abstract constant
(e.g. $\mathcal{L}_{syntax}(\synt{love}) = \lambda^{\circ} x y.\ x +
loves + y$). Managing this involves some tedious boilerplate that can be
avoided by another convenience facility. First, we define a lexicon that
produces the desired object term but with the object constant abstracted
out (e.g. $\lambda^{\circ} c x y.\ x + c + y$). Then, we can apply the
\texttt{lexicalizer} function to the object signature and to this
lexicon to get the desired lexicon which fills in the correct object
constants.

It is also useful to be able to easily insert the non-lexical constants
of the object signature into the object terms we are giving when
defining a lexicon. This service is provided by the anaphoric macro
\texttt{with-sig-consts} which introduces all of the non-lexical
constants of a signature into scope for easy use inside of lexicon
definitions.

Finally, since lexicons are again just relations, we can take their
unions and intersections using disjunction (\texttt{orr}) and
conjunction (\texttt{andr}), respectively.

@@

The crucial property that we need to check for in signatures is that the
relation between the identities of constants and their types is a
function. This simply reduces to querying the system for the number of
typed constants belonging to the signature and having a specific
identity. If this number is one for all the identities in our test set,
we have verified that (the tested subset of) the signature assigns
exactly one type to every one of its constants.

### 2013-06-26 outline.tex
Proofreading for publication

In the rest of this chapter, we proceed to introduce two grammatical
formalisms that are of interest to our work, the formalism of abstract
categorial grammars and the formalism interaction grammars. We talk
about a formal connection between the two and its value with respect to
constructing our grammar.

In Chapter~\ref{chap:implementation}, we present our software
implementation of the ACG machinery which will enable us to write and
test grammars lexicalized by a lexicon.

@@

We conclude our findings with Chapter~\ref{chap:conclusion}, in which we
summarize our findings and point out potential directions for future
work.

>>>>

In the rest of this chapter, we proceed to introduce two grammatical
formalisms that are of interest to our work, the formalism of abstract
categorial grammars and the formalism of interaction grammars. We talk
about a formal connection between the two and its value with respect to
constructing our grammar.

In Chapter~\ref{chap:implementation}, we present our software
implementation of the ACG machinery which will enable us to write and
test grammars lexicalized by Frilex.

@@

We conclude our report with Chapter~\ref{chap:conclusion}, in which we
summarize our findings and point out potential directions for future
work.
